---
title: "Classification of 1994 Census Income Data"
author: "Titus Njonjo"
date: "4/20/2020"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Problem Statement: 
To build a model that will predict if the income of any individual in the US is greater than or less than USD 50,000 based on the data available about that individual.
Data Set Description:
This Census Income dataset was collected by Barry Becker in 1994 and given to the public site http://archive.ics.uci.edu/ml/datasets/Census+Income. This data set will help you understand how the income of a person varies depending on various factors such as the education background, occupation, marital status, geography, age, number of working hours/week, etc.
PREDICTOR VARIABLES
-Age -Work-class -Final-weight -Education
-Education-num (Number of years of education)
-Marital-status -Occupation -Relationship -Race -Sex -Capital-gain -Capital-loss
-Hours-per-week -Native-country
RESPONSE VARIABLE
- Income-level

#Importing data
```{r}
trainFile = "adult.data"; testFile = "adult.test"
if (!file.exists (trainFile))
  download.file (url = "http://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data",
destfile = trainFile)
if(!file.exists(testFile))
  download.file (url = "http://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data",
destfile = testFile)

```
#reading data
```{r}
training <- read.table(trainFile,header=FALSE, sep=",", strip.white = TRUE,col.names = colNames, na.strings ="?", stringsAsFactors = TRUE)
training
colNames = c ("age", "workclass", "fnlwgt", "education",
"educationnum", "maritalstatus", "occupation",
"relationship", "race", "sex", "capitalgain",
"capitalloss", "hoursperweek", "nativecountry",
"incomelevel")
 
```
#display the structure of the data
```{r}
str(training)
```
#data cleaning,looking for NA values
```{r}
table(complete.cases(training))
summary(training[!complete.cases(training),])
```
#Removing missing values
```{r}
trainset = training[!is.na(training$workclass) & !is.na(training$occupation),]
trainset= training[!is.na(training$nativecountry),]
table(complete.cases(trainset))
trainset$fnlwgt=NULL
str(trainset)
```
  #data exploration
```{r}                       
 summary(trainset$age)
#boxplot for the age variable
boxplot(age~incomelevel, data= trainset,
        main="Imcome level versus the Age",
        xlab="Income level",
        ylab="Age",col="salmon")
#histogram for the income distribution
Below50K <- (trainset$incomelevel=="<=50k")
xlimit <- c(min(trainset$age),max(trainset$age))
ylimit <- c(0,1600)
hist1 <- qplot(trainset$age,data=trainset[Below50K,], margins=TRUE, binwidth=2, xlim=xlimit, ylim=ylimit, color= trainset$incomelevel)
hist2 <- qplot(trainset$age,data=trainset[!Below50K,], margins=TRUE, binwidth=2, xlim=xlimit, ylim=ylimit, color=trainset$incomelevel)
#grid_arrange(hist1,hist2 ,nrow=2)
#exploring the number of years in education
summary(trainset$educationnum)
boxplot(educationnum~incomelevel , data=trainset,
        main="Number of years in Education ",
        xlab="Income level",
        ylab="Education Num",
        col="salmon")
#Evaluate hours/week variable
summary (trainset$hoursperweek)
boxplot (hoursperweek ~ incomelevel, data = trainset,
main = "Hours Per Week distribution for different income levels",
xlab = "Income Levels", ylab = "Hours Per Week", col = "salmon")
```
```{r}
#Exploring work class variable
qplot(incomelevel, data=trainset,fill=workclass)+ facet_grid(.~workclass)
#exploring occupation variable
qplot(incomelevel, data=trainset, fill=occupation)+ facet_grid(.~occupation)
#exploring marital status
qplot(incomelevel, data=trainset, fill=maritalstatus)+facet_grid(.~maritalstatus)
#Exloring the relationship status
qplot(incomelevel, data=trainset, fill=relationship)+facet_grid(.~relationship)
#eploring the country native variable
#qplot(incomelevel, data=trainset, fill=nativecountry)+facet_grid(.~nativecountry)
```
#building model
```{r}
set.seed(30162)
trnctrl <- trainControl(method="cv", number=10) 
boostfit <- train(incomelevel ~ age + educationnum + relationship + workclass +
                    occupation + relationship + maritalstatus +
                    hoursperweek + capitalgain + capitalloss +
                    race + nativecountry,
                  trcontrol = trnctrl,
                  method="gbm", 
                  data=trainset, 
                  verbose=FALSE)
options(na.action = =na.warn)



```
```{r}
#Checking the accuracy of the model
#The output shows that our model calculates the income level of an individual with an accuracy of approximately 86%, which is a good number.
#So far, we used the training data set to build the model, now its time to validate the model by using the testing data set.###

confusionMatrix (trainSet$incomelevel, predict (boostFit, trainSet))


```

```{r}
#testing the model
testing <-  read.table (testFile, header = FALSE, sep = ",",
                        strip.white = TRUE, col.names = colNames,
                       na.strings = "?", fill = TRUE, stringsAsFactors = TRUE)

#structure  of the test data
str(testing)
#checking for missing values
table(complete.cases(testing))
summary(testing[!complete.cases(testing),])
#Removing NAs
TestSet = testing [!is.na (testing$workclass) & !is.na (testing$occupation), ]
TestSet = TestSet [!is.na (TestSet$nativecountry), ]
 
#Removing unnecessary variables
TestSet$fnlwgt = NULL
```
```{r}
#validating model
#Testing model
TestSet$predicted = predict (boostFit, TestSet)
table(TestSet$incomelevel, TestSet$predicted)
actuals_preds <- data.frame(cbind(actuals=TestSet$incomelevel, predicted=TestSet$predicted))
# make actuals_predicteds dataframe.
correlation_accuracy <- cor(actuals_preds)
head(actuals_preds)

```






